Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-08 Documents/Paper/ms.tex
tweaks

We will assume that the prior for $N$ is a discrete uniform distribution
between 0 and some $N_{\rm max}$, inclusive. The proposal starts by choosing
a new value $N'$ according to
\begin{eqnarray}
N' &=& \textnormal{mod}\left(N + \delta_N, N_{\rm max} + 1\right)
\end{eqnarray}
where $\delta_N$ is drawn from a heavy tailed
distribution which is a discrete analogue of Equation~\ref{eqn:heavytailed}.
The most probable values of $\delta_N$ are $\pm 1$, but values
of order $N_{\rm max}$ also have some probability, to allow fast exploration
when the target distribution is similar to the prior. If $N' = N$, $N'$ is
regenerated as $N+1$ with probability 0.5, and $N-1$ with probability 0.5,
and then wrapped back into the range $\{0,...,N_{\rm max}\}$ if necessary.

If $N' > N$, i.e. the proposal is to add components to the model,
the extra parameters needed, $\{\xx_i\}_{i=N+1}^{N'}$,
are drawn from their prior conditional on the current value of the
hyperparameters, i.e. the interim prior $p(\xx | \hyperparams)$.
If $N' < N$, i.e. the proposal is to remove components from the model,
then $N - N'$ components must be selected for removal. All
of the components have the same probability of being selected for removal.

@@

Let $F(x; \hyperparams)$ be a function that takes an component $x$ and transforms it
to a value $u$ that has a uniform distribution between 0 and 1, given $\hyperparams$.
If the component $x$ consists of a single scalar value, $F$ is the cumulative
distribution of the interim prior. Denote the inverse of $F$ by $G$.

@@

The ``dragging'' process works as follows, and must be carried out on
each component:
\begin{eqnarray}
u_i &:=& F(\xx_i; \hyperparams)\\
\xx_i' &:=& G(u_i; \hyperparams')
\end{eqnarray}
In other words, the components' parameters are transformed to $[0,1]$ using the
current value of the hyperparameters, and then transformed back using the
proposed value of the hyperparameters so they represent the new interim prior
rather than the old one.

@@

\subsection{Prior Distributions}
To infer $N$, the number of sinusoids, as well as all the periods, amplitudes,
and phases, we need to define prior distributions.
In the notation of Section~\ref{sec:introduction}, the parameters of the
each component are
\begin{eqnarray}
\xx_i &=& \{A_i, T_i, \phi_i\}.
\end{eqnarray}
For the interim prior $p(\xx_i | \hyperparams)$, we introduced a single
hyperparameter $\mu$, such that $A_i$ given $\mu$ has an exponential prior
with mean $\mu$. For the periods, we assigned a log-uniform distribution for
the periods between fixed boundaries, and a uniform distribution for the phases
between 0 and $2\pi$. In other words, our prior is only hierarchical for the
amplitudes $\{A_i\}$. The prior for $\mu$ was a log-uniform prior, with
probability density $\propto 1/\mu$, where $\mu \in [10^{-3}, 10^3]$.

@@

\subsection{Priors}
For the interim prior for the total fluxes of the galaxies, and for the
widths of the galaxies, we used the Pareto
distribution. The probability density function for a variable $x$ given
hyperparameters $x_{\rm min}$ and $a$ is
\begin{eqnarray}
p(x | x_{\rm min}, a) &=&
\left\{
\begin{array}{lr}
\frac{a x_{\rm min}^{a}}{x^{a + 1}}, & x \geq x_{\rm min}\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
Since we are using this for both the fluxes and the widths of the galaxies,
there will be four hyperparameters in total: two lower limits (one for the
fluxes and one for the widths) and two slopes. The lower limits were assigned
log-uniform priors between $10^{-3}$ and $10^3$, and uniform priors between 0
and 1 were assigned to the {\it inverses} of the slopes. This latter choice
was based on the fact that a Pareto distribution for $x$ is an exponential
distribution for $\ln(x)$, with scale length $1/a$.

The interim priors for the radius ratios $u$ and the flux ratios $w$ were both
uniform distributions. Since $u \in [0,1]$ and $v \in [0,1]$, the upper limits
of the uniform interim priors must be between 0 and 1. Therefore the prior
for the upper limit was a uniform distribution between 0 and 1, and the prior
for the lower limit given the upper limit was a uniform distribution between 0
and the upper limit.

>>>>

We will assume that the prior for $N$ is a discrete uniform distribution
between 0 and some $N_{\rm max}$, inclusive. The proposal starts by choosing
a new value $N'$ according to
\begin{eqnarray}
N' &=& \textnormal{mod}\left(N + \delta_N, N_{\rm max} + 1\right)
\end{eqnarray}
where $\delta_N$ is drawn from a heavy tailed
distribution which is a discrete analogue of Equation~\ref{eqn:heavytailed}
($N$ is treated as a real number in $[0, N_{\rm max} + 1]$ and then we take
the integer part at the end).
The most probable values of $\delta_N$ are $\pm 1$, but values
of order $N_{\rm max}$ also have some probability, to allow fast exploration
when the target distribution is similar to the prior. If $N' = N$, $N'$ is
regenerated as $N+1$ with probability 0.5, and $N-1$ with probability 0.5,
and then wrapped back into the range $\{0,...,N_{\rm max}\}$ if necessary.

If $N' > N$, i.e. the proposal is to add components to the model,
the extra parameters needed, $\{\xx_i\}_{i=N+1}^{N'}$,
are drawn from their prior conditional on the current value of the
hyperparameters, i.e. the conditional prior $p(\xx | \hyperparams)$.
If $N' < N$, i.e. the proposal is to remove components from the model,
then $N - N'$ components must be selected for removal. All
of the components have the same probability of being selected for removal.

@@

Let $F(x; \hyperparams)$ be a function that takes an component $x$ and transforms it
to a value $u$ that has a uniform distribution between 0 and 1, given $\hyperparams$.
If the component $x$ consists of a single scalar value, $F$ is the cumulative
distribution of the conditional prior. Denote the inverse of $F$ by $G$.

@@

The ``dragging'' process works as follows, and must be carried out on
each component:
\begin{eqnarray}
u_i &:=& F(\xx_i; \hyperparams)\\
\xx_i' &:=& G(u_i; \hyperparams')
\end{eqnarray}
In other words, the components' parameters are transformed to $[0,1]$ using the
current value of the hyperparameters, and then transformed back using the
proposed value of the hyperparameters so they represent the new conditional prior
rather than the old one.

@@

\subsection{Prior Distributions}
To infer $N$, the number of sinusoids, as well as all the periods, amplitudes,
and phases, we need to define prior distributions.
In the notation of Section~\ref{sec:introduction}, the parameters of the
each component are
\begin{eqnarray}
\xx_i &=& \{A_i, T_i, \phi_i\}.
\end{eqnarray}
For the conditional prior $p(\xx_i | \hyperparams)$, we introduced a single
hyperparameter $\mu$, such that $A_i$ given $\mu$ has an exponential prior
with mean $\mu$. For the periods, we assigned a log-uniform distribution for
the periods between fixed boundaries, and a uniform distribution for the phases
between 0 and $2\pi$. In other words, our prior is only hierarchical for the
amplitudes $\{A_i\}$. The prior for $\mu$ was a log-uniform prior, with
probability density $\propto 1/\mu$, where $\mu \in [10^{-3}, 10^3]$.

@@

\subsection{Priors}
For the conditional prior for the total fluxes of the galaxies, and for the
widths of the galaxies, we used the Pareto
distribution. The probability density function for a variable $x$ given
hyperparameters $x_{\rm min}$ and $a$ is
\begin{eqnarray}
p(x | x_{\rm min}, a) &=&
\left\{
\begin{array}{lr}
\frac{a x_{\rm min}^{a}}{x^{a + 1}}, & x \geq x_{\rm min}\\
0, & \textnormal{otherwise.}
\end{array}
\right.
\end{eqnarray}
Since we are using this for both the fluxes and the widths of the galaxies,
there will be four hyperparameters in total: two lower limits (one for the
fluxes and one for the widths) and two slopes. The lower limits were assigned
log-uniform priors between $10^{-3}$ and $10^3$, and uniform priors between 0
and 1 were assigned to the {\it inverses} of the slopes. This latter choice
was based on the fact that a Pareto distribution for $x$ is an exponential
distribution for $\ln(x)$, with scale length $1/a$. If we do not expect the
scale length to be very extreme then we might want to restrict it to $[0,1]$
with a uniform prior.

The conditional priors for the radius ratios $u$ and the flux ratios $w$ were both
uniform distributions. Since $u \in [0,1]$ and $v \in [0,1]$, the upper limits
of the uniform conditional priors must be between 0 and 1. Therefore the prior
for the upper limit was a uniform distribution between 0 and 1, and the prior
for the lower limit given the upper limit was a uniform distribution between 0
and the upper limit.
